Investigating magnetic dynamo evolution with TESS field dwarfs

### 2017-10-04 ms.tex
removing the authors

\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\RAVE}{{\it RAVE}}
\newcommand{\rave}{{\it RAVE}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

@@

\author{R. Angus\altaffilmark{\columbia}
        J. R. Davenport\altaffilmark{\ww}
        D. Foreman-Mackey\altaffilmark{\cca},
        S. Oh\altaffilmark{\princeton},
        D. Busazi\altaffilmark{\florida},
        A. Mann\altaffilmark{\columbia}
        D. Kipping\altaffilmark{\columbia}
}

\altaffiltext{\columbia}{Department of Astronomy, Columbia University, New
York, NY}
\altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}
\altaffiltext{\ww}{Western Washington University, Bellingham, WA}
\altaffiltext{\nsf}{NSF Fellow}
\altaffiltext{\florida}{Florida Gulf Coast University, Fort Myers, FL}
\altaffiltext{\cca}{Center for Computational Astronomy, Flatiron Institute,
New York, NY}
\altaffiltext{\hubble}{Hubble Fellow}
\altaffiltext{\princeton}{Princeton University, Princeton, NJ}

@@

% Age is a fundamental stellar parameter of great interest to at galactic
% archaeologists and exoplaneteers alike but unfortunately it is a difficult
% attribute to measure for main sequence F, G, K, and M stars in the field.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
Using precise light curves available from the \TESS\ mission, we expect to
measure the rotation periods of around ... stars.
Most of these rotation periods will be convertible into ages as they will be
intermediate age FGKM dwarfs.

Some outstanding puzzles regarding the nature of magnetic braking remain to be
answered.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars.
{\bf We propose to help answer these questions by providing a catalog of \TESS\
rotation periods for all FFI and CTL stars.}
We also intend to use gyrochronology in the cases where it {\it is} applicable
to learn about the evolution of exoplanets and the galaxy as a whole.

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars, as shown in Figure
\ref{fig:davenport} \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI and CTL targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a age discontinuous age
distribution, the feature should be local; it could
disappear at greater distances or along different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI and CTL targets to the upcoming data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
With the April 2018 data release from Gaia (DR2), we estimate that we will be
able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3
kpc.

@@

Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods for both FFI and
two-minute cadence stars, with the maximum measurable period strongly
depending on the length of time it is observed for.
Of particular interest are the, potentially long, rotation periods measurable
for stars that fall in the continuous viewing zones.
Many of these will have spectra from the \TESS\/HERMES survey, from which it
will be possible to calculate isochronal ages, enabling further calibration of
the gyrochronology relations.
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics with \Gaia\.
Many stars in the TESS CTL will have proper motions, parallaxes, positions and
radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
% PI Angus is currently using the \kepler\ rotation periods of stars in the
% first \Gaia\ data release to calibrate the relations between rotation period
% and vertical action dispersion, two tracers of age.
% Since this relation is likely vary with both galactic longitude and latitude,
% the rotation periods produced from \TESS\ light curves will enable a
% comprehensive, all-sky calibration of these relations.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$.

In addition, we will use the rotation periods of comoving stars identified in
the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.
We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
Of these, 34 stars in 17 pairs are already in the CTL list.
Because the stars in these pairs are assumed to be recently disrupted
binaries, we expect them to have the same age and measuring their rotation
periods will therefore allow us to test gyrochronology.

Finally, we hope to use the improved gyrochronology relations, calibrated
using \TESS\ data to infer the dependence of planet frequency on age and
position in the galaxy.
\kepler\ revealed tantalizing hints that exoplanets on short orbital periods
have larger radii and are more infrequent around young stars \citep{mann2017a,
mann2017b, rizzuto2017}.
Do these trends continue into older ages and do they exist in the field?
We will answer these questions by inferring ages for \TESS\ field stars.

@@

\section{Expected Impact}
We will provide light curves and rotation periods for both the two-minute
cadence and FFI targets.

>>>>

\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\Wfirst}{{\it Wfirst}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Gaia}{{\it Gaia}}
\newcommand{\gaia}{{\it Gaia}}
\newcommand{\RAVE}{{\it RAVE}}
\newcommand{\rave}{{\it RAVE}}
\newcommand{\hermes}{{\it HERMES}}
\newcommand{\HERMES}{{\it HERMES}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

@@

% \author{R. Angus\altaffilmark{\columbia}
%         J. R. Davenport\altaffilmark{\ww}
%         D. Foreman-Mackey\altaffilmark{\cca},
%         S. Oh\altaffilmark{\princeton},
%         D. Busazi\altaffilmark{\florida},
%         A. Mann\altaffilmark{\columbia}
%         D. Kipping\altaffilmark{\columbia}
% }

% \altaffiltext{\columbia}{Department of Astronomy, Columbia University, New
% York, NY}
% \altaffiltext{\simons}{Simons Fellow, RuthAngus@gmail.com}
% \altaffiltext{\ww}{Western Washington University, Bellingham, WA}
% \altaffiltext{\nsf}{NSF Fellow}
% \altaffiltext{\florida}{Florida Gulf Coast University, Fort Myers, FL}
% \altaffiltext{\cca}{Center for Computational Astronomy, Flatiron Institute,
% New York, NY}
% \altaffiltext{\hubble}{Hubble Fellow}
% \altaffiltext{\princeton}{Princeton University, Princeton, NJ}

@@

% Age is a fundamental stellar parameter of great interest to at galactic
% archaeologists and exoplaneteers alike but unfortunately it is a difficult
% attribute to measure for main sequence F, G, K, and M stars in the field.
Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age, and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star’s rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star’s surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\%) quasi-periodic changes in the stellar brightness.
Using precise light curves available from the \TESS\ mission, we expect to
measure the rotation periods of around ... stars.
For many of these stars it will be possible to infer an age from their
rotation periods via gyrochronology.

Some outstanding mysteries regarding the nature of magnetic braking have been
revealed, but thus-far unanswered by, \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars.
{\bf We propose to help answer these questions by providing a catalog of \TESS\
rotation periods for all FFI and CTL stars.}
We also intend to use gyrochronology in the cases where it {\it is} applicable
to learn about the evolution of exoplanets across the galaxy.

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport discovered this period bimodality extends throughout
all masses in the Kepler rotation sample for nearby stars
\citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The TESS FFI and CTL targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the TESS FFI and CTL targets to the upcoming data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
With the April 2018 data release from Gaia (DR2), we estimate that we will be
able to study rotation periods for G dwarfs in the TESS FFIs out to $\sim$ 3
kpc.

@@

Although the classical spin-down law of \citet{skumanich1972}: Period
$\propto$ Age$^1/2$ holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods for both FFI and
two-minute cadence stars, with the maximum measurable period strongly
depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the continuous viewing zones.
Many of these will have spectra from the \TESS\ /\HERMES\ survey, from which it
will be possible to calculate isochronal ages, enabling further calibration of
the gyrochronology relations.
We will also use the rotation periods of TESS field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many stars in the TESS CTL will have proper motions, parallaxes, positions and
radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action angles of the stars, both of which are age indicators.
% PI Angus is currently using the \kepler\ rotation periods of stars in the
% first \Gaia\ data release to calibrate the relations between rotation period
% and vertical action dispersion, two tracers of age.
% Since this relation is likely vary with both galactic longitude and latitude,
% the rotation periods produced from \TESS\ light curves will enable a
% comprehensive, all-sky calibration of these relations.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.

In addition, we will use the rotation periods of comoving stars identified in
the first \Gaia\ data release \citep{oh2016} to quantify the accuracy and
precision of gyrochronology.
We have identified 38 stars in 19 pairs with \Gaia\ proper motions and radial
velocities from \RAVE\ that are observable by \TESS, using {\tt tvguide}.
Of these, 34 stars in 17 pairs are already in the CTL list.
Because the stars in these pairs are assumed to be recently disrupted
binaries, we expect them to have the same age and measuring their rotation
periods will therefore provide a test of gyrochronology.

Finally, we hope to use the improved gyrochronology relations, calibrated
using \TESS\ data to infer the dependence of planet frequency on age and
position in the galaxy.
\kepler\ revealed tantalizing hints that exoplanets on short orbital periods
have larger radii and are more infrequent around young stars \citep{mann2017a,
mann2017b, rizzuto2017}.
Do these trends continue into older ages and do they exist in the field?
We will answer these questions by measuring rotation periods for all \TESS\
field stars.

@@

\section{Expected Impact}
We will provide rotation periods for all two-minute cadence and FFI targets.
